---
title: "Raw Facts, 2018"
author: "Charlotte Mack"
date: "`r format(Sys.Date())`"
output: github_document
---

A set of comparisons for the CPS Enrollment data, to be visualized or arranged in tables elsewhere. At date, "first" year is 2006 and "last" year is 2018.

```{r options}
options(digits = 2)
```


```{r libraries}
library(tidyverse)
```

```{r data}
enrollment_all_hs <- read_rds("./enrollment_all_hs.Rds")
```


```{r}
# Filter first and this year from data
first_last <- enrollment_all_hs %>% filter(year %in% c(2006, 2018))
```

```{r}
# Counts of schools, total and by governance
first_last %>% 
    group_by(year) %>%  
    count()

first_last %>% 
    group_by(year, govern) %>%  
    count()
```

```{r}
# Enrollments, total and by governance
all <- first_last %>%  
    group_by(year) %>%  
    summarize(all_enrollment = sum(total_hs))
# (Match column 2 of this into the next one, then calculate the percentages.)

first_last %>%  
    group_by(year, govern) %>%  
    summarize(enrollment = sum(total_hs)) 
```


```{r}
# Largest and smallest enrollments overall and by governance
# 

# Largest overall:
first_last %>% 
    filter(year == 2006) %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018) %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)
    
# Smallest overall:
first_last %>% 
    filter(year == 2006) %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018) %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)


# Largest of regular governance:
first_last %>% 
    filter(year == 2006 & govern == "regular") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "regular") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)
    
# Smallest of regular governance:
first_last %>% 
    filter(year == 2006 & govern == "regular") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "regular") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

# Largest of charter governance:
first_last %>% 
    filter(year == 2006 & govern == "charter") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "charter") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)
    
# Smallest of charter governance:
first_last %>% 
    filter(year == 2006 & govern == "charter") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "charter") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

```

```{r births and deaths}
# Using anti-joins, determine which schools (by school_id) left the database
# during the period and which ones entered:
first <- first_last %>% filter(year == 2006)
this <- first_last %>% filter(year == 2018)

# School IDs present in 2006 but not in 2018:
changes_f <- anti_join(first, this, by = "school_id")
changes_f %>% glimpse()

# School IDs not present in 2006 but appearing in 2018:
changes_t <- anti_join(this, first, by = "school_id")
changes_t %>% glimpse()

```


```{r growth numbers}
# Form a data subset having only schools represented in both years, by id, and
# use this to calculate enrollment changes, i.e. absolute, percent, and
# annualized rate, over the period.

deltas <- first_last %>% 
    filter(!(school_id %in% changes_f$school_id) & 
               !(school_id %in% changes_t$school_id)) %>% 
    mutate(delta = ifelse(year == 2018, 
                          total_hs - lag(total_hs, 
                                         1, 
                                         order_by = school_id), 
                          NA), 
           delta_pct_by_100 = delta/lag(total_hs, 
                                 1, 
                                 order_by = school_id),
           delta_rate = ((1 + delta_pct_by_100)^(1/12)) - 1) %>% 
    filter(year == 2018)




# Highest growth over period:
deltas %>%  
    arrange(desc(delta)) %>%  
    select(common_name, total_hs, delta) %>%  
    head(10)

# Greatest decline over period:
deltas %>%  
    arrange(delta) %>%  
    select(common_name, total_hs, delta) %>%  
    head(87)
```

```{r growth percent}
# Highest growth over period by percent:
deltas %>%  
    arrange(desc(delta_pct_by_100)) %>%  
    select(common_name, delta, delta_pct_by_100) %>%  
    head(10)

# Greatest decline over period by percent:
deltas %>%  
    arrange(delta_pct_by_100) %>%  
    select(common_name, delta, delta_pct_by_100) %>%  
    head(25)
```
```{r annualized growth rates}
# Fastest rate of growth over period, annualized:
deltas %>%  
    arrange(desc(delta_rate)) %>%  
    select(common_name, delta_pct_by_100, delta_rate) %>%  
    head(10)

# Fastest rate of decline over period, annualized
deltas %>%  
    arrange(delta_rate) %>%  
    select(common_name, delta_pct_by_100, delta_rate) %>%  
    head(25)

```

```{r}
# Prepare a complete case data.frame using school_id as the only key variable:
both <- first_last %>%  
    filter(!(school_id %in% changes_f$school_id) &
               !(school_id %in% changes_t$school_id)) %>%  
    select(govern, school_id, year, total_hs)
# NB: There are three school_id values that apply to more than one common_name.
# This causes spurious NA to appear when the spread operation above is applied
# to a both data.frame that includes the common_name variable. The three schools
# are Westinghouse/Westinghouse II (609693), Perspectives Charter (400064) which
# is probably a typo error in the database, and Dugan/Peace and Education
# (610386). One way around this until I decide whether to alter the main
# database is to use the deltas data.frame to correct the names and then take
# the ranks:

ranked <- both %>% 
    mutate(common_name = deltas$common_name[match(school_id,
                                                  deltas$school_id)]) %>% 
    spread(year, total_hs, drop = TRUE) %>% 
    rename(hs_06 = `2006`, hs_18 = `2018`) %>% 
    mutate(rank_06 = trunc(rank(-hs_06, ties.method = "min")), 
           rank_18 = trunc(rank(-hs_18, ties.method = "min")))

ranked %>%  
    arrange(rank_06) %>%  
    select(govern, common_name, rank_06, rank_18) #  %>% print(15)

ranked %>%  
    arrange(desc(rank_06)) %>%  
    select(govern, common_name, rank_06, rank_18) #  %>% print(15)
```

Use dumbbells with color aesthetic for year, arrow connecting 06 to 18, for a portion of the data --- maybe above the median --- or in panels for the entire set. Write up the factsheet as a demonstration of things one can do with data reshaping and grammar of graphics.