---
title: "Raw Facts, 2018"
author: "Charlotte Mack"
date: "`r format(Sys.Date())`"
output: github_document
---

A set of comparisons for the CPS Enrollment data, to be visualized or arranged in tables elsewhere. At date, "first" year is 2006 and "last" year is 2018.

```{r options}
options(digits = 2)
```


```{r libraries}
library(tidyverse)
```

```{r data}
enrollment_all_hs <- read_rds("./data/enrollment_all_hs.Rds")
```


```{r}
# Filter first and this year from data
first_last <- enrollment_all_hs %>% filter(year %in% c(2006, 2018))
```

```{r }
# Summary statistics, total and by governance
first_last %>%
    filter(year == 2006) %>%
    summary()
```

```{r}
# Counts of schools, total and by governance
first_last %>% 
    group_by(year) %>%  
    count()

first_last %>% 
    group_by(year, govern) %>%  
    count()
```

```{r}
# # Enrollments, total and by governance
# all <- first_last %>%  
#     group_by(year) %>%  
#     summarize(all_enrollment = sum(total_hs))
# # (Match column 2 of this into the next one, then calculate the percentages.)
# 
# first_last %>%  
#     group_by(year, govern) %>%  
#     summarize(enrollment = sum(total_hs),
#                                pct =
#                                    enrollment/all$all_enrollment)
```
                                  
Proportions of schools by governance type:
    
```{r }
first_last %>%
    mutate(year = as.factor(year)) %>%
    ggplot(aes(x = year, fill = fct_infreq(govern))) +
    geom_bar(position = position_stack(reverse = TRUE), 
             width = .2) +
    coord_flip() +
    scale_fill_manual(values = c("darkblue", "green4", "yellow")) +
    theme(legend.position = "bottom",
          legend.spacing.x = unit(.2, "cm"),
          legend.title = element_blank()) 
```


```{r}
# Largest and smallest enrollments overall and by governance
# 

# Largest overall:
first_last %>% 
    filter(year == 2006) %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018) %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)
    
# Smallest overall:
first_last %>% 
    filter(year == 2006) %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018) %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)


# Largest of regular governance:
first_last %>% 
    filter(year == 2006 & govern == "regular") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "regular") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)
    
# Smallest of regular governance:
first_last %>% 
    filter(year == 2006 & govern == "regular") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "regular") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

# Largest of charter governance:
first_last %>% 
    filter(year == 2006 & govern == "charter") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "charter") %>%  
    arrange(desc(total_hs)) %>%  
    select(govern, common_name, total_hs)
    
# Smallest of charter governance:
first_last %>% 
    filter(year == 2006 & govern == "charter") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

first_last %>% 
    filter(year == 2018 & govern == "charter") %>%  
    arrange(total_hs) %>%  
    select(govern, common_name, total_hs)

```

```{r births and deaths}
# Using anti-joins, determine which schools (by school_id) left the database
# during the period and which ones entered:
first <- first_last %>% filter(year == 2006)
this <- first_last %>% filter(year == 2018)

# School IDs present in 2006 but not in 2018:
changes_f <- anti_join(first, this, by = "school_id")
changes_f %>% glimpse()

# School IDs not present in 2006 but appearing in 2018:
changes_t <- anti_join(this, first, by = "school_id")
changes_t %>% glimpse()

```


```{r growth numbers}
# Form a data subset having only schools represented in both years, by id, and
# use this to calculate enrollment changes, i.e. absolute, percent, and
# annualized rate, over the period.

deltas <- first_last %>% 
    filter(!(school_id %in% changes_f$school_id) & 
               !(school_id %in% changes_t$school_id)) %>% 
    mutate(delta = ifelse(year == 2018, 
                          total_hs - lag(total_hs, 
                                         1, 
                                         order_by = school_id), 
                          NA), 
           delta_pct_by_100 = delta/lag(total_hs, 
                                 1, 
                                 order_by = school_id),
           delta_rate = ((1 + delta_pct_by_100)^(1/12)) - 1) %>% 
    filter(year == 2018)




# Highest growth over period:
deltas %>%  
    arrange(desc(delta)) %>%  
    select(common_name, total_hs, delta) %>%  
    head(10)

# Greatest decline over period:
deltas %>%  
    arrange(delta) %>%  
    select(common_name, total_hs, delta) %>%  
    head(87)
```

```{r growth percent}
# Highest growth over period by percent:
deltas %>%  
    arrange(desc(delta_pct_by_100)) %>%  
    select(common_name, delta, delta_pct_by_100) %>%  
    head(10)

# Greatest decline over period by percent:
deltas %>%  
    arrange(delta_pct_by_100) %>%  
    select(common_name, delta, delta_pct_by_100) %>%  
    head(25)
```

Twenty-four schools lost more than half their enrollment over the data period:

```{r }
deltas %>% 
    filter(delta_pct_by_100 <= -.50) %>% 
    ggplot(aes(x = fct_rev(fct_reorder(common_name, delta_pct_by_100)),
               y = delta_pct_by_100)) +
    geom_bar(stat = "Identity", width = .25) +
    coord_flip()
```

Here are the enrollment timelines for those schools:
```{r }

decliners <- deltas %>% 
    filter(delta_pct_by_100 <= -.50) %>% 
    select(common_name)

enrollment_all_hs %>% 
    filter(common_name %in% decliners$common_name) %>%
    ggplot(aes(x = year, y = total_hs)) +
    geom_line() +
    facet_wrap(~ common_name, nrow = 6)

```

```{r annualized growth rates}
# Fastest rate of growth over period, annualized:
deltas %>%  
    arrange(desc(delta_rate)) %>%  
    select(common_name, delta_pct_by_100, delta_rate) %>%  
    head(10)

# Fastest rate of decline over period, annualized
deltas %>%  
    arrange(delta_rate) %>%  
    select(common_name, delta_pct_by_100, delta_rate) %>%  
    head(25)

```

```{r}
# Prepare a complete case data.frame using school_id as the only key variable:
both <- first_last %>%  
    filter(!(school_id %in% changes_f$school_id) &
               !(school_id %in% changes_t$school_id)) %>%  
    select(govern, school_id, year, total_hs)
# NB: There are three school_id values that apply to more than one common_name.
# This causes spurious NA to appear when the spread operation above is applied
# to a both data.frame that includes the common_name variable. The three schools
# are Westinghouse/Westinghouse II (609693), Perspectives Charter (400064) which
# is probably a typo error in the database, and Dugan/Peace and Education
# (610386). One way around this until I decide whether to alter the main
# database is to use the deltas data.frame to correct the names and then take
# the ranks:

ranked <- both %>% 
    mutate(common_name = deltas$common_name[match(school_id,
                                                  deltas$school_id)]) %>% 
    spread(year, total_hs, drop = TRUE) %>% 
    rename(hs_06 = `2006`, hs_18 = `2018`) %>% 
    mutate(rank_06 = trunc(rank(-hs_06, ties.method = "min")), 
           rank_18 = trunc(rank(-hs_18, ties.method = "min")))

ranked %>%  
    arrange(rank_06) %>%  
    select(govern, common_name, rank_06, rank_18) #  %>% print(15)

ranked %>%  
    arrange(desc(rank_06)) %>%  
    select(govern, common_name, rank_06, rank_18) #  %>% print(15)
```

The next graph shows high schools ranked by total grade 9 to 12 enrollment, anchored in the 2006-2007 school year. The direction of the arrows shows whether a school moved up (to the left) or down in the ranking. Schools that made large moves up the ranking had been originally among the smaller schools, no surprise, and of the three most striking cases, two were schools that early in the data period were being re-opened and phased in after extensive physical makeover (Lindblom and Westinghouse). The larger number of above-median schools that fell to below median, or even first quartile, ranks lacks such simple organic explanation.

```{r ranking dumbells}
ranked %>% 
    mutate(common_name = 
               fct_reorder(common_name, rank_06, median)) %>%
    select(-starts_with("hs")) %>% 
    gather(key = rank_yr, 
           value = ranking, 
           -govern, 
           -school_id, 
           -common_name) %>% 
    ggplot(aes(x = fct_rev(common_name), 
               y = ranking, 
               color = rank_yr)) + 
    geom_point(position = position_jitter(width = 0, height = 0.6)) +
    coord_flip() + 
    geom_path(aes(group = common_name), 
              lineend = "butt", 
              linejoin = "mitre", 
              arrow = grid::arrow(angle = 15, 
                                  length = unit(.2, "cm"), 
                                  ends = "last", 
                                  type = "closed"))

```